The Space of Feasible Execution Times for Asynchronous Periodic Task Systems using Definitive Idle Times

### 2013-09-30 dit-paper_ieee.tex
Less than 6 pages. You are now closed #18

\author{
\IEEEauthorblockN{Thomas Chapeaux and Paul Rodriguez}
\IEEEauthorblockA{Universit\'{e} Libre de Bruxelles / ECE Paris\\
tchapeau@ulb.ac.be, paurodri@ulb.ac.be}
\and
\IEEEauthorblockN{Laurent George}
\IEEEauthorblockA{University of Paris-Est / ECE Paris\\
lgeorge@ieee.org}
\and
\IEEEauthorblockN{Jo\"el Goossens}
\IEEEauthorblockA{Universit\'{e} Libre de Bruxelles\\
joel.goossens@ulb.ac.be}
}

@@

  Even then, the number of constraints to consider is exponential in the number
  of tasks. While it has been shown that constraints that cover intervals ending
  after the end of the first busy period are also redundant, this value depends on the $C_i$ and is
  therefore not usable for the purposes of this paper. The
  following concept, initially described by \cite{lipariaverage}, is thus introduced. It gives an upper bound for the end of the first busy period and does not depend on the execution times.

@@

  For synchronous systems in which the execution times are not known, the first DIT $t_d$
  is our ``best guess'' at an upper limit on the end of the busy period. As it was sufficient
  to check the DBF test in intervals $[0, t]$ before the end of the first busy period. This time interval is WCETs dependant. A more general feasility condition valid for any WCET is as follows:
  \begin{equation}
  		\dbf{0,t} \leqslant t \; \forall t \leqslant t_d
  		\iff \text{The system is feasible}
  \end{equation}

@@

	\subsection{Solving the modular system}
		A sketch of the construction algorithm to compute $t_d$ can be found in
		Algorithm~\ref{alg:algoCRP} (note that $x^{-1} \; (mod \; y)$ means the
		modular inverse of $x$ in the naturals modulo $y$). The congruences must
		first be modified in order to obtain pairwise coprime moduli. Each $n_i$
		is factored into prime powers $p^{k_1}_1 p^{k_2}_2 \cdots p^{k_m}_m$
		with $k_j$ the highest integer s.t. $p^{k_j}_j$ divides $n_i \; \forall j
		\; \mid 1 \leqslant j \leqslant m$. Each original constraint $t_d \equiv x_i
		\; (mod \; n_i)$ is replaced by the equivalent set \cite{knuth1969art}
		\begin{equation*}
			\begin{split}
				t_d \equiv x_i 	& \; (mod \; p^{k_1}_1) \\
				t_d \equiv x_i 	& \; (mod \; p^{k_2}_2) \\
								& \vdots \\
				t_d \equiv x_i 	& \; (mod \; p^{k_m}_m) \\
			\end{split}
		\end{equation*}
		This operation might create
		duplicate constraints, which are removed.
		As shown in \cite{knuth1969art}, the resulting system can be solved using
		Algorithm~\ref{alg:algoCRP}. For ease of reading, we call $C$ the set of
		resulting constraints. We define two functions on the set of
		constraints. Assuming contraint $c$ of the form $t_d \equiv x \; (mod \; p)$,
		we have $p(c) = p$ and $x(c) = x$

@@

		Algorithm~\ref{alg:algoCRP} finds one DIT given one combination of $a_i$
		values. However using it in practice is very inefficient as there are
		exponentially many different combinations of $a_i$ values in the number of
		tasks (the cartesian product of all the $A_i$ sets), and resolving
		exponentially many CRP instances is prohibitively slow.
		In this paper a faster method for finding the FPDIT is proposed. It consists
		in merging all these CRP instances into one problem (finding all DITs), which can be solved faster by
		Algorithm~\ref{alg:multiACRP}, as a significant part of the intermediate
		results are shared between instances. First, the same $C$ set is obtained,
		but instead of one $x_i$ value per task, all variants in $X_i$ spawn a set of
		constraints. All the duplicate and useless constraints are removed. Then we
		define $P$ the set of all powers of prime numbers appearing as moduli in the
		congruences (the $p_j^{k_j}$ values). For all $p \in P$, we define $C_p$ the set of
		constraints where $p$ is the modulo.
		\begin{algorithm}
			\caption{Modified Gauss Algorithm}
			\label{alg:multiACRP}
			\begin{algorithmic}[1]
                \REQUIRE $P$, $C_{p}$, $x(c)$: as defined above
				\REQUIRE $H$ is the hyperperiod of the system

@@

				\STATE $results \leftarrow \text{empty list}$
				\STATE append $0$ to $results$
				\FORALL{$p \in P$}
					\STATE $newresults \leftarrow \text{empty list}$
					\FORALL{$chunk \in sumchunks[p]$}
						\FORALL{$r \in results$}
							\STATE $\text{append} \; (r + chunk) \; \text{to} \; newresults$
						\ENDFOR
					\ENDFOR
					\STATE $results \leftarrow newresults$
				\ENDFOR
				\RETURN $r \; (mod \; H)$ for all $r \in results$
			\end{algorithmic}
		\end{algorithm}
		In the first loop over $P$ of Algorithm~\ref{alg:multiACRP}, all the
		required intermediate values are computed. In the second loop, those
		intermediate values are added in every possible way. Both
		Algorithm~\ref{alg:algoCRP} and Algorithm~\ref{alg:multiACRP} require the
		computation of one modular inversion per constraint, which
		is done in a linear number of modular inversions per task. However, to
		find all DITs Algorithm~\ref{alg:algoCRP} has to be executed once for each possible
		combination of $a_i$ values, which is exponential in the number of tasks.
		Algorithm~\ref{alg:multiACRP} has the same theoretical complexity as the
		computation of all DITs still requires an exponential number of answers in the
		number of tasks, but it is nonetheless significantly faster in practice. One
		drawback of Algorithm~\ref{alg:multiACRP} is that the replacement of each
		constraint by sets of constraints with powers of primes as moduli is not
		equivalent. Algorithm~\ref{alg:multiACRP} might ouput values that are not
		DITs. However, as we only need the FPDIT, it remains practical to check those
		solutions one by one from the lowest to the highest, and stop as soon as the
		first correct result is found.

@@

	We thus checked the existence condition on 10000 systems generated by:
	\begin{itemize}
		\item Generating utilization values with the \texttt{UUniFast} algorithm described in \cite{bini2005measuring}, with:
		\begin{itemize}
			\item System utilization chosen uniformly between 0.25 and 0.75
			\item The number of tasks set to several values (see Fig.~\ref{fig:noFPDIT})
		\end{itemize}
		\item Choosing the $T_i$ uniformly between 5 and 20 (in order to have a manageable value of $H$), which also gives us the $C_i = \lfloor U_i \, T_i \rfloor$
		\item Choosing the $O_i$ with a normal distribution of mean $T_{min}$ and standard deviation $\frac{T_{max} - T_{min}}{2}$
		\item Choosing the $D_i$ uniformly in the interval $[T_i - CDF(T_i - C_i), T_i]$, where $CDF$ varies between 0 and 1 and is called the \emph{Constrained Deadline Factor}. It allows us to generate systems where the deadlines are closer to the period.
	\end{itemize}

	The number of tasks and the constrained Deadline Factor is shown to have a direct influence on the density of FPDIT. Indeed, Fig~\ref{fig:noFPDIT} show the percentage of systems having no FPDIT for different values of CDF.

@@

			As shown in Fig. \ref{fig:asyncBehavior}, the system begins in the
			\emph{incomplete period}, where all tasks are not in the system. Then, after
			$O_{max}$ time units, every tasks is in the system and the \emph{transient
			period} begins. Then, after $O_{max} + H$ time units, the system is faced
			with the same pattern of arrival as in $O_{max}$. However, because in
			$O_{max} + H$ every task is in the system and in $O_{max}$ some were missing,
			the state of the system will be different. The \emph{stationary period} then
			begins. At instant $O_{max} + 2H$ it is guaranteed that the system is in the
			exact same situation it was at instant $O_{max} + H$.

			\begin{figure}[h]
				\[
					\begin{array}{r||c|c|c|l}
									& \text{Incomplete} & \text{Transient}	& \text{Stationary} & \cdots \\
						\text{Name} & \text{period} 	& \text{period} 	& \text{period}  	& \\ \hline
						\text{size} & O_{max} 			& H 				& H 				& \cdots \\ \hline
						t \geqslant & 0 				& O_{max} 			& O_{max} + H 		& \cdots
					\end{array}
				\]
				\begin{center}
				\caption{Behavior of an asynchronous system under EDF.}
				\label{fig:asyncBehavior}
				\end{center}
			\end{figure}

			For this reason, as shown by \cite{leung1982complexity}, it is sufficient (but pessimistic) to
			check for
			intervals included in the interval $[O_{max}, O_{max} + 2H]$, which has an exponential length
			in the number of tasks. A better interval is given in \cite{choquet2004minimal} using the
			notion of cyclic idle times (denoted $t_c$), which are dependent of the execution
			time and therefore cannot be used directly to characterize the C-Space. Their interval is $[0, t_c + H]$ and is valid for
			every cyclic idle time $t_c$.

			In the following section, we extend the latter interval using the previously defined notion
			of periodic DIT, a particular case of cyclic idle times which does not depend on the execution
			times.

@@

The FPDIT, if it exists, must occur in the transient period. Indeed, it cannot occur in the incomplete period by definition ($t_d > O_{max}$) and if it occurred in the $k^{th}$ stationary period, another DIT should have occurred in the transient period at instant $t_d - k H$ (a contradiction).

As the FPDIT happens when all tasks are in the system, we know that the system will arrive in the exact same state one hyperperiod later ($t_d + H$) (A rigorous proof is given in \cite{choquet2004minimal} for any cyclic idle time). We can thus restrict the DBF test to intervals included in the interval $[t_d, t_d + H]$ if the FPDIT exists. This will always be a better interval than $[O_{max}, O_{max} + 2H]$ and if the first periodic DIT does not exist (which can be verified as explained in Section \ref{sct:FPDITexist}), the latter interval can be used.

@@

The FPDIT happens at $t=15$ and we have $H = 15$. The study interval using the FPDIT is thus $[15, 30]$, which contains 11 intervals $[a,d]$ with $a$ being the arrival of a job, $d$ being the deadline of a job, and $15 \leqslant a < d \leqslant 30$. Those intervals are
$[15, 17]$, $[15, 22]$, $[15, 27]$, $[15, 30]$, $[20, 22]$, $[20, 27]$, $[20, 30]$, $[23, 27]$, $[23, 30]$, $[25, 27]$ and $[25, 30]$. For comparison, the original study interval of $[O_{max}, O_{max} + 2H]$ is $[8, 38]$ which yields 57 intervals.

@@

We now apply Algorithm \ref{alg:prunRedun} to remove redundancies, leaving us with the following constraints:
	\begin{equation}
		\left\{
			\begin{array}{ccccc}
				& & C_2 & \leqslant & 2 \\
				C_1 & + & C_2 & \leqslant & 7
			\end{array}
		\right.
	\end{equation}
The C-space thus contains 11 points $(C_1, C_2) = (1, 1)$, $(2, 1)$, $(3, 1)$, $(4, 1)$, $(5, 1)$, $(6, 1)$, $(1, 2)$, $(2, 2)$, $(3, 2)$, $(4, 2)$ and $(5, 2)$.

Let us now consider the same system but with a synchronous pattern of arrival ($O_1 = O_2 = 0$). The first DIT happens at time $t_d = 7$. The intervals to consider are thus of the form $[0, t]$ where $t$ is a deadline happening before or at time $t_d$. This yields two intervals ($[0, 2]$ and $[0, 7]$) and the following constraints describing the C-space, none of them being redundant:
$$
\left\{
  \begin{array}{ccccc}
    & & C_2 & \leqslant & 2 \\
    C_1 & + & 2 C_2 & \leqslant & 7
  \end{array}
\right.
$$
This time, the C-space only contains 8 points: $(1, 1)$, $(2, 1)$, $(3, 1)$, $(4, 1)$, $(5, 1)$, $(1, 2)$, $(2, 2)$, $(3, 2)$.

Fig. \ref{fig:cspaceComp} shows the two C-spaces in the plane. The dark gray region is the C-space of the synchronous system and the light gray region is the surface gain given by the asynchronous system.

\begin{figure}[h]
\begin{center}
  \includegraphics[width=0.3\textwidth]{figs/cspace_example.png}
  \caption{C-space of an asynchronous system and its corresponding synchronous system}
  \label{fig:cspaceComp}
\end{center}
\end{figure}

>>>>

\author{
\IEEEauthorblockN{Thomas Chapeaux and Paul Rodriguez}
\IEEEauthorblockA{Universit\'{e} Libre de Bruxelles / ECE Paris\\
 Avenue Franklin Roosevelt 50\\
 1050 Brussels, Belgium\\
tchapeau@ulb.ac.be, paurodri@ulb.ac.be}
\and
\IEEEauthorblockN{Laurent George}
\IEEEauthorblockA{University of Paris-Est / ECE Paris\\
Quai de Grenelle, 37\\
75015 Paris\\
lgeorge@ieee.org}
\and
\IEEEauthorblockN{Jo\"el Goossens}
\IEEEauthorblockA{Universit\'{e} Libre de Bruxelles\\
 Avenue Franklin Roosevelt 50\\
 1050 Brussels, Belgium\\
joel.goossens@ulb.ac.be}
}

@@

  Even then, the number of constraints to consider is exponential in the number
  of tasks. While it has been shown \cite{baruah1990algorithms} that constraints that cover intervals ending
  after the end of the first busy period are also redundant, this value depends on the $C_i$ and is
  therefore not usable for the purposes of this paper. The
  following concept, initially described by \cite{lipariaverage}, is thus introduced. It gives an upper bound for the end of the first busy period and does not depend on the execution times.

@@

  For synchronous systems in which the execution times are not known, the first DIT $t_d$
  is an upper limit on the end of the busy period regardless of the WCET values. As it was sufficient
  to check the DBF test in intervals $[0, t]$ before the end of the first busy period, a condition valid for any WCET is as follows:
  \begin{equation}
  		\dbf{0,t} \leqslant t \; \forall t \leqslant t_d
  		\iff \text{The system is feasible}
  \end{equation}

@@

	\subsection{Solving the modular system}
		A sketch of the construction algorithm to compute $t_d$ can be found in
		Algorithm~\ref{alg:algoCRP} (note that $x^{-1} \; (mod \; y)$ means the
		modular inverse of $x$ in the naturals modulo $y$). The congruences must
		first be modified in order to obtain pairwise coprime moduli. Each $n_i$
		is factored into prime powers $p^{k_1}_1 p^{k_2}_2 \cdots p^{k_m}_m$
		with $k_j$ the highest integer s.t. $p^{k_j}_j$ divides $n_i \; \forall j
		\; \mid 1 \leqslant j \leqslant m$. Each original constraint $t_d \equiv x_i
		\; (mod \; n_i)$ is replaced by the following equivalent set of constraints \cite{knuth1969art}:
		\begin{equation*}
			\begin{split}
				t_d \equiv x_i 	& \; (mod \; p^{k_1}_1) \\
				t_d \equiv x_i 	& \; (mod \; p^{k_2}_2) \\
								& \vdots \\
				t_d \equiv x_i 	& \; (mod \; p^{k_m}_m) \\
			\end{split}
		\end{equation*}
		This operation might create
		duplicate constraints, which are removed. We call $C$ the set of
		resulting constraints. We define two functions on the set of
		constraints, assuming contraint $c$ of the form $t_d \equiv x \; (mod \; p)$
		we have $p(c) = p$ and $x(c) = x$. As shown in \cite{knuth1969art}, the resulting system can be solved
        using Algorithm~\ref{alg:algoCRP}.

@@

		Algorithm~\ref{alg:algoCRP} finds one DIT given one combination of $a_i$
		values. However using it in practice is very inefficient as there are
		exponentially many different combinations of $a_i$ values in the number of
		tasks (the cartesian product of all the $A_i$ sets), and resolving
		exponentially many CRP instances is prohibitively slow.

		We propose a faster method for finding the FPDIT which consists
		in merging all these CRP instances into one problem (finding all DITs), which can be solved faster by
		Algorithm~\ref{alg:multiACRP}, as a significant part of the intermediate
		results are shared between instances. First, the same $C$ set is obtained,
		but instead of one $x_i$ value per task, all variants in $X_i$ spawn a set of
		constraints. All duplicate and useless constraints are removed. We then
		define $P$ the set of all powers of prime numbers appearing as moduli in the
		congruences (the $p_j^{k_j}$ values). For all $p \in P$, we define $C_p$ the set of
		constraints where $p$ is the modulo.

		\begin{algorithm}
			\caption{Modified Gauss Algorithm}
			\label{alg:multiACRP}
			\begin{algorithmic}[1]
                \REQUIRE $P$, $C_{p}$, $x(c)$: as defined above
				\REQUIRE $H$ is the hyperperiod of the system

@@

				\STATE $results \leftarrow \text{empty list}$
				\STATE append $0$ to $results$
				\FORALL{$p \in P$}
					\STATE $newresults \leftarrow \text{empty list}$
					\FORALL{$chunk \in sumchunks[p]$}
						\FORALL{$r \in results$}
							\STATE $\text{append} \; (r + chunk) \; \text{to} \; newresults$
						\ENDFOR
					\ENDFOR
					\STATE $results \leftarrow newresults$
				\ENDFOR
				\RETURN $r \; (mod \; H)$ for all $r \in results$
			\end{algorithmic}
		\end{algorithm}
		In the first loop over $P$ of Algorithm~\ref{alg:multiACRP}, all the
		required intermediate values are computed. In the second loop, those
		intermediate values are added in every possible way. Both
		Algorithm~\ref{alg:algoCRP} and Algorithm~\ref{alg:multiACRP} require the
		computation of one modular inversion per constraint, which
		is done in a linear number of modular inversions per task. However, to
		find all DITs Algorithm~\ref{alg:algoCRP} has to be executed once for each possible
		combination of $a_i$ values, which is exponential in the number of tasks.
		Algorithm~\ref{alg:multiACRP} has the same theoretical complexity as the
		computation of all DITs still requires an exponential number of answers in the
		number of tasks, but it is nonetheless significantly faster in practice.

        One drawback of Algorithm~\ref{alg:multiACRP} is that the replacement of each
		constraint by sets of constraints with powers of primes as moduli is not
		equivalent. Algorithm~\ref{alg:multiACRP} might output values that are not
		DITs. However, as we only need the FPDIT, it remains practical to check those
		solutions one by one from the lowest to the highest, and stop as soon as the
		first correct result is found.

@@

	We thus checked the existence condition on 10000 systems generated by:
	\begin{itemize}
		\item Generating utilization values with the \texttt{UUniFast} algorithm described in \cite{bini2005measuring}, with:
		\begin{itemize}
			\item System utilization chosen uniformly between 0.25 and 0.75
			\item The number of tasks set to several values (see Fig.~\ref{fig:noFPDIT})
		\end{itemize}
		\item Choosing the $T_i$ uniformly between 5 and 20 (in order to have a manageable value of $H$), which also gives us the $C_i = \lfloor U_i \, T_i \rfloor$
		\item Choosing the $O_i$ with a normal distribution of mean $T_{min}$ and standard deviation $\frac{T_{max} - T_{min}}{2}$
		\item Choosing the $D_i$ uniformly in the interval \mbox{$[T_i - CDF \cdot (T_i - C_i), T_i]$}, where $CDF$ varies between 0 and 1 and is called the \emph{Constrained Deadline Factor}. It allows us to generate systems where the deadlines are closer to the period.
	\end{itemize}

	The number of tasks and the Constrained Deadline Factor is shown to have a direct influence on the density of FPDIT. Indeed, Fig~\ref{fig:noFPDIT} show the percentage of systems having no FPDIT for different values of CDF.

@@

			The system begins in the
			\emph{incomplete period}, where all tasks are not in the system. Then, after
			$O_{max}$ time units, every task has arrived in the system and the \emph{transient
			period} begins. Then, after $O_{max} + H$ time units, the system is faced
			with the same pattern of arrival as in $O_{max}$. However, because in
			$O_{max} + H$ every task is in the system and in $O_{max}$ some were missing,
			the state of the system may differ. The \emph{stationary period} then
			begins. At instant $O_{max} + 2H$ it is guaranteed that the system is in the
			exact same situation it was at instant $O_{max} + H$.

			% For this reason, as shown by \cite{leung1982complexity}, it is sufficient (but pessimistic) to
			As shown in \cite{leung1982complexity}, it is sufficient to
            apply the demand-bound function test in
			intervals included in the interval $[O_{max}, O_{max} + 2H]$, which has an exponential length
			in the number of tasks. A better interval is given in \cite{choquet2004minimal} using the
			notion of cyclic idle times (denoted $t_c$), which are idle times occurring at fixed intervals of
            length $H$ and whose value depends of the tasks execution
			time, therefore they cannot be used directly to characterize the C-Space. Their interval is
            $[0, t_c + H]$ and is valid for every cyclic idle time $t_c$.

			In the following section, we show how the latter interval can be used with the previously defined
            notion of periodic DIT, a particular case of cyclic idle times which does not depend on the execution
			times.

@@

The FPDIT, if it exists, must occur after in the transient period. Indeed, $t_d > O_{max}$ by definition, and if it occurred in the $k^{th}$ stationary period, another DIT should have occurred in the transient period at instant $t_d - k H$ (a contradiction).

Despite $t_d$ being in the theoretical transient period, the system will be in the same state at instant $t_d + H$ (A rigorous proof is given in \cite{choquet2004minimal} for any cyclic idle time). We can thus restrict the DBF test to intervals included in the interval $[t_d, t_d + H]$ if the FPDIT exists. This will always be a smaller interval than $[O_{max}, O_{max} + 2H]$ and if the first periodic DIT does not exist (which can be verified as explained in Section \ref{sct:FPDITexist}), the latter interval can be used.

@@

The FPDIT happens at $t=15$ and we have $H = 15$. The study interval using the FPDIT is thus $[15, 30]$, which contains 11 intervals $[a,d]$ with $a$ being the arrival of a job, $d$ being the deadline of a job, and $15 \leqslant a < d \leqslant 30$. For comparison, the original study interval of $[O_{max}, O_{max} + 2H]$ is $[8, 38]$ which yields 57 intervals.

@@

We now apply Algorithm \ref{alg:prunRedun} to remove redundancies, leaving us with the following constraints:
	\begin{equation}
		\left\{
			\begin{array}{ccccc}
				& & C_2 & \leqslant & 2 \\
				C_1 & + & C_2 & \leqslant & 7
			\end{array}
		\right.
	\end{equation}

Let us now consider the same system but with a synchronous pattern of arrival ($O_1 = O_2 = 0$). The first DIT happens at time $t_d = 7$. The intervals to consider are thus of the form $[0, t]$ where $t$ is a deadline happening before or at time $t_d$. This yields two intervals ($[0, 2]$ and $[0, 7]$) and the following constraints describing the C-space, none of them being redundant:
$$
\left\{
  \begin{array}{ccccc}
    & & C_2 & \leqslant & 2 \\
    C_1 & + & 2 C_2 & \leqslant & 7
  \end{array}
\right.
$$

We see that the offsets allow three more points to be in the C-space ($(C_1, C_2) = (4, 2), (5, 2), (6, 1)$).

% \begin{figure}[h]
% \begin{center}
%   \includegraphics[width=0.3\textwidth]{figs/cspace_example.png}
%   \caption{C-space of an asynchronous system and its corresponding synchronous system}
%   \label{fig:cspaceComp}
% \end{center}
% \end{figure}
